Detailed disease progression of 88 patients hospitalized with Covid-19 in the Czech Republic: An exploratory analysis

### 2020-09-17 manuscript/manuscript.Rmd
ACP score - basic treatment

---
title: "Detailed disease progression of 88 patients hospitalized with Covid-19 in the Czech Republic: An exploratory analysis" 
output: 
  #rticles::arxiv_article
  #pdf_document
  bookdown::pdf_document2: 
     toc: false
  # bookdown::pdf_book: 
  #    base_format: rticles::arxiv_article
abstract: "We collected a multi-centric retrospective dataset of patients who were admitted to N hospitals in Czech Republic and tested positive for SARS-CoV-2. The dataset contains baseline patient characteristics, breathing support required, pharmacological treatment received and multiple markers on daily resolution. To assess association between treatments, markers and patient outcomes we performed multiverse analysis, observing how the conclusions change between defensible choices of statistical model, predictors included in the model and other analytical degrees of freedom."
bibliography: Covid19.json
---

@@

@@

First class of models are frequentist multistate models under the proportional hazards assumption as implemented in the `coxph` function from the `survival` package [@http://zotero.org/users/5567156/items/I9UVTUJF]. We primarily use a model with competing risks for death and discharge from hospital (see Figure \@ref(fig:modelstates)a).

Second class of models are Bayesian hidden Markov models of disease progression implemented via a custom extension to the `brms` package [@http://zotero.org/users/5567156/items/7ER3ZSER]. The parametrization of the HMM is inspired by [@http://zotero.org/users/5567156/items/F9K2L5Q8] where an underlying continous process with given rates only between the neighbouring state and the transition matrix across all states is then computed as the total probability of transition across all possible paths. This class of models does not satisfy the proportional hazards assumption, instead, it is assumed the process has the Markov property - i.e. that the unobserved state and the covariates at a given day contain all the information to determine probabilities of the states on the next day. We primarily use a model that explicitly uses all of the recorded levels of breathing support and has a hidden (unobserved) element whether the patient is improving or worsening. The states and available transitions are shown in Figure \@ref(fig:modelstates)b.

@@

@@

In Figure \@ref(fig:progression) we show the disease progression for all patients.

@@

@@

The results of the multiverse analysis for association between drugs used and outcomes is shown in Figure \@ref(fig:multiverseresultsdrugs). In the current dataset, we see a possibly strong negative association between using Hydroxychloroquine and risk of death across all models, although for several models we cannot rule out that the association is of negligible strength. Across all models, the data offer no definitive conclusions on the association between taking Hydroxychloroquine and length of hospital stay (represented as risk for being discharged). After adjusting for Hydroxychloroquine usage, we also cannot make any strong claims on the association of using Azithromycin with either outcome. In all cases, the results seem to not be strongly affected by model choice.

We additionally investigated the association of several markers with disease progression. Due to low number of patients in the current dataset, the results are inconclusive and heavily dependent on model choice. All of the coefficient estimates can be seen in Figure \@ref(fig:multiverseresultsmarkers)

>>>>

---
title: "Detailed disease progression of 88 patients hospitalized with Covid-19 in the Czech Republic: An exploratory analysis" 
output: 
  #rticles::arxiv_article
  #pdf_document
  bookdown::pdf_document2: 
     toc: false
  # bookdown::pdf_book: 
  #    base_format: rticles::arxiv_article
abstract: "We collected a multi-centric retrospective dataset of patients who were admitted to N hospitals in Czech Republic and tested positive for SARS-CoV-2. The dataset contains baseline patient characteristics, breathing support required, pharmacological treatment received and multiple markers on daily resolution. To assess association between treatments, markers and patient outcomes we performed multiverse analysis, observing how the conclusions change between defensible choices of statistical model, predictors included in the model and other analytical degrees of freedom. Additionally we performed external validation of several proposed prognostic models for Covid-19 severity."
bibliography: Covid19.json
date: "Version: `r format(Sys.time(), '%Y-%m-%d')`"
---

@@

Additionally, many methods for predicting disease severity of Covid-19 were published, but the methods are at high risk of bias and lack external validation [@http://zotero.org/users/5567156/items/TS57HJ3T].

@@

First class of models are frequentist multistate models under the proportional hazards assumption as implemented in the `coxph` function from the `survival` package [@http://zotero.org/users/5567156/items/I9UVTUJF]. We primarily use a model with competing risks for death and discharge from hospital (see Figure&nbsp;\@ref(fig:modelstates)a).

Second class of models are Bayesian hidden Markov models of disease progression implemented via a custom extension to the `brms` package [@http://zotero.org/users/5567156/items/7ER3ZSER]. The parametrization of the HMM is inspired by [@http://zotero.org/users/5567156/items/F9K2L5Q8] where an underlying continous process with given rates only between the neighbouring state and the transition matrix across all states is then computed as the total probability of transition across all possible paths. This class of models does not satisfy the proportional hazards assumption, instead, it is assumed the process has the Markov property - i.e. that the unobserved state and the covariates at a given day contain all the information to determine probabilities of the states on the next day. We primarily use a model that explicitly uses all of the recorded levels of breathing support and has a hidden (unobserved) element whether the patient is improving or worsening. The states and available transitions are shown in Figure&nbsp;\@ref(fig:modelstates)b.

@@

## Evaluating prognostic models

We searched the living systematic review of Covid-19 prognostic models [@http://zotero.org/users/5567156/items/TS57HJ3T] for models that could be applied to our dataset (i.e. where we have gathered all the input features). This resulted in five usable models: [@http://zotero.org/users/5567156/items/EN4GICJ6; @http://zotero.org/users/5567156/items/INZVABQI; @http://zotero.org/users/5567156/items/3A2RN4ZS; @http://zotero.org/users/5567156/items/4DDUVEKS; @http://zotero.org/users/5567156/items/X4GWJIP5].

@@

In Figure&nbsp;\@ref(fig:progression) we show the disease progression for all patients.

@@

## Association between treatment, markers and outcomes

@@

The results of the multiverse analysis for association between drugs used and outcomes is shown in Figure&nbsp;\@ref(fig:multiverseresultsdrugs). In the current dataset, we see a possibly strong negative association between using Hydroxychloroquine and risk of death across all models, although for several models we cannot rule out that the association is of negligible strength. Across all models, the data offer no definitive conclusions on the association between taking Hydroxychloroquine and length of hospital stay (represented as risk for being discharged). After adjusting for Hydroxychloroquine usage, we also cannot make any strong claims on the association of using Azithromycin with either outcome. In all cases, the results seem to not be strongly affected by model choice.

We additionally investigated the association of several markers with disease progression. Due to low number of patients in the current dataset, the results are inconclusive and heavily dependent on model choice. All of the coefficient estimates can be seen in Figure&nbsp;\@ref(fig:multiverseresultsmarkers)

## Evaluating prediction models

The ACP index [@http://zotero.org/users/5567156/items/EN4GICJ6] is a simple score that categorizes the patients into 3 categories based on age and C-reactive protein (CRP): grade 1 (age $<$ 60 years and CRP $<$ 34 mg/L), grade 2 (age $\geq$ 60
years and CRP $<$ 34 mg/L or age $<$ 60 years and CRP $\geq$ 34 mg/L), grade 3 (age $\geq$ 60 years and CRP $\geq$ 34 mg/L) and was evaluated primarily on 12-day mortality. Despite the score being developed on patients in Wuhan, China, the 12-day mortality in our dataset is in rough agreement with the values in the original manuscript as shown in Table \@ref(tab:acpresults).

```{r acpresults, echo=FALSE}
acp_data <- wide %>%  
  group_by(patient_id) %>%
  filter(any(!is.na(CRP))) %>%
  summarise(
    first_CRP_day = min(day[!is.na(CRP)]),
    first_CRP = CRP[day == first_CRP_day],
    worst_breathing_12 = max(breathing[day >= first_CRP_day & day <= first_CRP_day + 12]),
    .groups = "drop") %>% 
  inner_join(data$patient_data, by = c("patient_id")) %>%
  filter(outcome %in% c("Death","Discharged") | last_record - first_CRP_day >= 12) %>%
  mutate(    
    alive_12 = last_record > first_CRP_day + 12 | outcome == "Discharged",
    ACP_grade = case_when(
       age < 60 & first_CRP < 34 ~ 1,
       age >= 60 & first_CRP >= 34 ~ 3,
       TRUE ~ 2
    ) 
  )

acp_data %>%
  group_by(ACP_grade) %>%
  summarise(`patients` = n(), `12-day mortality` = scales::percent(mean(!alive_12), accuracy = 0.1), .groups = "drop") %>%
  arrange(ACP_grade) %>%
  mutate(`Li et al. 95% CI` = c("0%", "0 - 11.3%", "19.8 - 44.3%")) %>%
  rename(`ACP grade` = ACP_grade) %>%
  knitr::kable(booktabs = TRUE, caption = "Comparing the 12-day mortality based on ACP score (where available) with the 95\\% confidence intervals (CI) for 12-day mortality reported by Lu et al.")

# acp_data %>%
#   group_by(ACP_grade, worst_breathing_12) %>%
#   summarise(mean(alive_12))
```

(TODO: the other prediction models)
